Testing Charts: viewers' perceptual accuracy in surveys

### 2023-07-29 SDSS-2023/testing-charts-accuracy.tex
a few wording edits from KR full editing pass-through

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{lscape}
\author[1]{Kiegan Rice\thanks{Corresponding author. Email: rice-kiegan@norc.org}}
\author[2]{Heike Hofmann\footnote{Email: hofmann@iastate.edu}}
\author[1]{Nola du Toit\footnote{Email: dutoit-nola@norc.org}}
\author[1]{Edward Mulrow\footnote{Email: mulrow-edward@norc.org}}
\affil[1]{NORC at the University of Chicago}
\affil[2]{Department of Statistics, Iowa State University}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Testing Charts: viewers' perceptual accuracy in surveys},
  pdfkeywords={list, keywords, here, 3-6, are, fine},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\maketitle
\begin{abstract}
The use of visuals is a key component in scientific communication.
Decisions about the design of a data visualization should be informed by
what design elements best support the audience's ability to perceive and
understand the components of the data visualization. We build on the
foundations of Cleveland and McGill's work in graphical perception,
employing a large, nationally-representative, probability-based panel of
survey respondents to test perception in statistical charts. Our
findings provide actionable guidance for data visualization
practitioners to employ in their work.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

What do viewers see when we show them a data chart? A data chart -- at
its core -- maps values to graphical elements: quantitative elements are
represented as measurable features, such as position, size, or shade.
Modern data visualizations are much more than a simple, objective
mapping of values to a plane; they contain contextual and design
elements, and are often structured to support the viewer in
understanding a particular view of a set of data or specific pattern
underlying the values. Structural choices, such as the choice of a
graphical element (e.g., bar or pie), orientation of the chart, and
sizing and relative positioning of elements within the chart, are
determined by the data visualization practitioner. The design of a data
visualization impacts a viewer's ability to achieve that understanding;
a poorly designed data visualization may leave viewers struggling to
understand the content or context, or make it difficult to complete
accurate and useful comparisons of values across groups or time points.
More broadly, the design of a data visualization can change how viewers
interact with the chart.

The viewer's employment of comparisons of the components is a crucial
step in the process of interacting with and understanding of chart.
Cleveland and McGill (1984) observed as such, and in their seminal study
defined the better visual among a pair as the one that allows viewers to
make more accurate comparisons. Based on mappings of quantitative
variables to different graphical elements, Cleveland and McGill's study
resulted in a ranking of perceptual tasks (such as ranking the larger of
two lengths or areas) from most accurate to least accurate, which was
then extended by Mackinlay (1986) to a theoretical framework ranking
tasks' order along their ordinal and nominal scales.

@@

\autoref{fig-tasks}(a,d) shows the two stacked bar charts shown to
participants in the first task. The marked tiles in each plot are 155
pixels apart, which leads to a JND of 3.5 pixels based on Lu et al.
(2022)`s model. The heights of the bars are 205 (left) and 213 pixels
(right), corresponding to about twice the JND. This difference should
lead to a relatively high accuracy rate for participants and
simultaneously limit the amount of frustration resulting from a task
that is perceived as 'too hard'.

Both charts show the same data with slight modifications to the order of
the levels -- the first and second level in each of the bars are
reversed between the left and the right chart. Both charts are displayed
at the same size, i.e.~in both cases both the difference in size between
the bars and the horizontal distance between the bars is the exact same
amount. This leaves the vertical positioning of the bars as the only
difference between the charts. Any differences in observed responses can
therefore be attributed to this difference in presentation.

@@

The visual mappings in the second task, shown in \autoref{fig-tasks}(b,
e), are identical to the first task, but the axes of the chart are
rotated so that the stacked bars are represented in a horizontal format.
This represents a structural change in how the data are presented to the
viewer while preserving the pixel size of the elements viewers must
compare. Tasks 2 and 3 were asked within the same survey round (Round 3)
and the full sample was randomly split among respondents, with 50\% of
participants seeing the Task 2 stimuli and 50\% of participants seeing
the Task 3 stimuli.

@@

The images utilized in the third task, \autoref{fig-tasks}(c,f), again
represent the same data as in the first two tasks, and the overall image
has dimensions which are identical to the images in the first two tasks,
displaying at the same size during the survey. The length of bars is
increased to fit the new dimensions. This increases the difference in
the length of bars from previously 8 pixels to 13 pixels. The widths of
the tiles are adjusted accordingly (from 50 to 30 pixels) to keep the
overall area of the tiles approximately constant.

@@

\begin{itemize}
\item
  A is bigger
\item
  B is bigger
\item
  They are the same
\end{itemize}

@@

Participants were recruited as part of NORC's AmeriSpeak
panel\footnote{NORC is a non-partisan organization that collects data
  for government and non-government clients}, which utilizes a
probability-based sampling methodology and samples U.S. households from
NORC's National Sample Frame that provides coverage of over 97\% of U.S.
households. The current panel size is 54,001 panel members aged 13 and
over residing in over 43,000 households (Dennis 2019, updated 2022).
Each test was conducted using the AmeriSpeak Omnibus survey, which runs
biweekly and samples around 1,000 U.S. adults to answer questions on a
variety of topics. Our tests require visual inspection of an image; for
this reason, our survey questions were only presented to web-based
panelists and not panelists who respond via phone interviews. The panel
has 45,565 web-based panelists, representing 93.3\% of the total panel
and 96.3\% of the total panel weights.

@@

When combining responses from different surveys, weights are rescaled
with respect to the total population, so that their order after
combining still reflects their relative importance.

We \emph{combine} (rather than cumulate) a set of \(\ell\) surveys
\(S_1\), \(S_2\), \ldots{} \(S_\ell\) (with \$ \ell \ge 2\$ ) as
described in O'Muircheartaigh and Pedlow (2002), by multiplying weights
in \(S_i\) by \(\lambda_i\), for \(1 \le i \le \ell\).
\(\lambda_i \in [0,1]\) with \(\sum_i \lambda_i = 1\) is given as

@@

The data for this paper were collected in several rounds as part of the
NORC Omnibus. The resulting number of participants, effective sample
sizes, and \(\lambda\) values are shown in \autoref{tbl-rounds}.

\hypertarget{tbl-rounds}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
\caption{\label{tbl-rounds}Survey rounds: dates, number of participants
(nominal sample size), effective sample size, sum of weights, and
factors for the combining surveys as discussed above.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Date
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\# Participants
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Effective Sample Size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Sum of weights \(\sum_i w_i\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(\lambda_i\)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Date
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\# Participants
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Effective Sample Size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Sum of weights \(\sum_i w_i\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(\lambda_i\)
\end{minipage} \\
\midrule()
\endhead
Round 1 & April 2022 & 933 & 521.1 & 934.9 & \(\lambda_1\) = 0.343 \\
Round 2 & May 2022 & 953 & 485.7 & 953.4 & \(\lambda_2\) = 0.320 \\
Round 3 & June 2022 & 921 & 513.5 & 923.1 & \(\lambda_3\) = 0.338 \\
\textbf{Total} & --- & 2807 & 1520.8 & & \\
\bottomrule()
\end{longtable}

@@

\caption{\label{fig-alignment}On the left (a), a stacked barchart shows
the number of respondents with correct (green) and incorrect (grey)
responses to the two comparison questions. When tiles are aligned along
the same axis, more than twice the number of responses are correct. The
shaded area along the top of the green tiles corresponds to 95\%
confidence intervals around (marginal) correct responses. On the right
(b), a parallel coordinate plot shows all combinations of responses.
There's a huge asymmetry in the number of responses where participants
answered only one of the questions correctly. A lot more responses are
correct when comparing aligned tiles than unaligned tiles.}

@@

where \(\ell \in \{1, 2\}\); \(i \in \{1, 2\}\) is the comparison type
(1 = Aligned, 2 = Unaligned), and \(j \in \{1, 2, 3\}\) is the chart
design, with 1 = Vertical, 2 = Horizontal, and 3 = Horizontal wide. The
estimated values and 95\% confidence intervals are shown in
Table~\ref{tbl-rounds-123}. The lowercase letters indicate significances
between pairs of estimates: two estimates (in the same column) are
significantly different at 5\% level, if they do not have any letter in
common (Piepho 2004).

\hypertarget{tbl-rounds-123}{}
\begin{longtable}{lrrlrrl}
\caption{\label{tbl-rounds-123}Odds for the cell-means model, letters behind numbers indicate pairwise
significances. Within the same column values are significantly different
(at 5\%) if they do not share the same letter. }\tabularnewline

\caption*{
{\large Odds of accuracy by task and chart type}
} \\ 
\toprule
 & \multicolumn{3}{c}{correct | same or wrong} & \multicolumn{3}{c}{correct or same | wrong} \\ 
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
 & Est. &           [95\% CI] &   & Est. &           [95\% CI] &   \\ 
\midrule
\multicolumn{7}{l}{Unaligned} \\ 
\midrule
Horizontal & $0.22$ &  [$0.15$, $0.32$] & a & $5.54$ &  [$4.04$, $7.59$] & a \\ 
Horizontal wide & $0.26$ &  [$0.19$, $0.37$] & ab & $6.20$ &  [$4.46$, $8.61$] & a \\ 
Vertical & $0.41$ &  [$0.36$, $0.47$] & b & $6.90$ &  [$5.75$, $8.28$] & a \\ 
\midrule
\multicolumn{7}{l}{Aligned} \\ 
\midrule
Horizontal & $0.63$ &  [$0.49$, $0.81$] & c & $14.02$ &  [$9.23$, $21.30$] & b \\ 
Horizontal wide & $2.67$ &  [$2.04$, $3.48$] & d & $17.12$ &  [$10.49$, $27.95$] & b \\ 
Vertical & $1.51$ &  [$1.33$, $1.71$] & e & $11.33$ &  [$8.99$, $14.28$] & b \\ 
\bottomrule
\end{longtable}

@@

Using linear scores for the response of Certainty, with `not certain at
all' assigned a score of 1 and `extremely certain' assigned a score of
5, we can estimate the effects of task, chart design, and correctness on
certainty by using a cell-means model of the form:

@@

where \(k = 1, ..., N\), \(\mu_{ij\ell(k)}\) is average certainty
(measured on a scale from 1 to 5) of the four combinations of task and
correctness by each of the three designs, where \(i = 1, 2\) encodes
unaligned/aligned, and \(j=1,2\) encodes wrong, correct, and
\(\ell = 1, 2, 3\) encodes horizontal, wide horizontal, and vertical
stacked bar charts, respectively. We also assume that errors are
normally distributed,
i.e.~\(\epsilon_k \stackrel{i.i.d}{\sim} N(0, \sigma^2)\) for all
\(k = 1, ..., N\). The results are shown in Figure~\ref{fig-certainty}.
What we find, is that the highest scores for certainty are associated
with comparing aligned tiles. Certainty scores are not significantly
different between correct and wrong responses. Scores are (mostly)
significantly lower for the unaligned task. Interestingly, the lowest
scores of certainty are associated with correct responses, here.

@@

\caption{\label{fig-certainty}Certainty by task, chart design, and
correctness. Correct answers in the unaligned task have the lowest
certainty scores associated with them. Incorrect responses on the
unaligned task show a significant boost in certainty in the vertical
stacked bar. For horizontal stacked bars the difference between aligned
and unaligned charts matters significantly. Errorbars around the points
show estimated scores (on a scale of 1 to 5) with corresponding 95\% CI.
The letters underneath the bars indicate significances. The scores for
two bars are significantly different if they do not share a letter.}

@@

\caption{\label{fig-demographics}Panelists' demographics matter,
particularly, when the task difficulty increases. For aligned tiles,
gender, age, and education are not significant factors. However, income
levels do have a (small) effect. When income levels increase, the
percentage of wrong answers decreases, while the percentage of correct
answers increases slightly (significant at below 0.05). For the more
difficult task of comparing unaligned tiles, demographics are more
significant. Higher levels of education and higher levels of income are
associated with a significant increase of panelists choosing a response
of `they are the same', resulting in significant decreases of both
correct answers and wrong answers.}

>>>>

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{lscape}
\author[1]{Kiegan Rice\thanks{Corresponding author. Email: rice-kiegan@norc.org}}
\author[2]{Heike Hofmann\footnote{Email: hofmann@iastate.edu}}
\author[1]{Nola du Toit\footnote{Email: dutoit-nola@norc.org}}
\author[1]{Edward Mulrow\footnote{Email: mulrow-edward@norc.org}}
\affil[1]{NORC at the University of Chicago}
\affil[2]{Department of Statistics, Iowa State University}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Testing Charts: viewers' perceptual accuracy in surveys},
  pdfkeywords={graphical perception, data visualization, survey data},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\maketitle
\begin{abstract}
The use of visuals is a key component in scientific communication.
Decisions about the design of a data visualization should be informed by
what design elements best support the audience's ability to perceive and
understand the components of the data visualization. We build on the
foundations of Cleveland and McGill's work in graphical perception,
employing a large, nationally-representative, probability-based panel of
survey respondents to test perception in statistical charts. Our
findings provide actionable guidance for data visualization
practitioners to employ in their work.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

What do viewers see when we show them a data chart? A data chart -- at
its core -- maps values to graphical elements: quantitative elements are
represented as measurable features, such as position, size, or shade.
Modern data visualizations are much more than a simple, objective
mapping of values to a plane; they contain contextual and design
elements, and are often structured to support the viewer in
understanding a particular view of a set of data or specific pattern
underlying the values. Structural choices, such as the choice of a
graphical element (e.g., bar or pie), orientation of the chart, and
sizing and relative positioning of elements within the chart, are
determined by the data visualization practitioner. The design of a data
visualization impacts a viewer's ability to achieve the intended
understanding; a poorly designed data visualization may leave viewers
struggling to understand the content or context, or make it difficult to
complete accurate and useful comparisons of values across groups or time
points. More broadly, the design of a data visualization can change how
viewers interact with the chart.

The viewer's employment of comparisons of the components is a crucial
step in the process of interacting with and understanding the chart.
Cleveland and McGill (1984) observed as such, and in their seminal study
defined the better visual among a pair as the one that allows viewers to
make more accurate comparisons. Based on mappings of quantitative
variables to different graphical elements, Cleveland and McGill's study
resulted in a ranking of perceptual tasks (such as identifying the
larger of two lengths or areas) from most accurate to least accurate,
which was then extended by Mackinlay (1986) to a theoretical framework
ranking tasks' order along their ordinal and nominal scales.

@@

\autoref{fig-tasks}(a,d) shows the two stacked bar charts shown to
participants in the first task. The marked tiles in each plot are 155
pixels apart, which leads to a JND of 3.5 pixels based on Lu et al.
(2022)`s model. The heights of the bars are 205 (left) and 213 pixels
(right), corresponding to about twice the JND. This difference should
lead to a relatively high accuracy rate for participants and
simultaneously limit the amount of frustration resulting from a task
that is perceived as `too hard'.

Both charts show the same data with slight modifications to the order of
the levels -- the first and second level in each of the bars are
reversed between the chart in (a) and the chart in (d). Both charts are
displayed at the same size, i.e.~in both cases both the difference in
size between the bars and the horizontal distance between the bars is
the exact same amount. This leaves the vertical positioning of the bars
as the only difference between the charts. Any differences in observed
responses can therefore be attributed to this difference in
presentation.

@@

The visual mappings in the second task, shown in
\autoref{fig-tasks}(b,e), are identical to the first task, but the axes
of the chart are rotated so that the stacked bars are represented in a
horizontal format. This represents a structural change in how the data
are presented to the viewer while preserving the pixel size of the
elements viewers must compare. Tasks 2 and 3 were asked within the same
survey round (Round 3) and the full sample was randomly split among
respondents, with 50\% of participants seeing the Task 2 stimuli and
50\% of participants seeing the Task 3 stimuli.

@@

The images utilized in the third task, \autoref{fig-tasks}(c,f), again
represent the same data as in the first two tasks, and the overall image
has dimensions which are identical to the images in the first two tasks,
displaying at the same size during the survey. However, the aspect ratio
is adjusted to make the plotting area match that of Task 1, with the
mapping area being wider rather than taller. The length of bars is
increased to fit the new dimensions. This increases the difference in
the length of bars to 13 pixels (previously 8 pixels). The widths of the
tiles are adjusted accordingly (from 50 to 30 pixels) to keep the
overall area of the tiles approximately constant.

@@

\begin{itemize}
\item
  A is bigger
\item
  B is bigger
\item
  They are the same''
\end{itemize}

@@

Participants were recruited as part of NORC's AmeriSpeak
panel\footnote{NORC is a non-partisan organization that collects data
  for government and non-government clients.}, which utilizes a
probability-based sampling methodology and samples U.S. households from
NORC's National Sample Frame that provides coverage of over 97\% of U.S.
households. The current panel size is 54,001 panel members aged 13 and
over residing in over 43,000 households (Dennis 2019, updated 2022).
Each test was conducted using the AmeriSpeak Omnibus survey, which runs
biweekly and samples around 1,000 U.S. adults to answer questions on a
variety of topics. Our tests require visual inspection of an image; for
this reason, our survey questions were only presented to web-based
panelists and not panelists who respond via phone interviews. The panel
has 45,565 web-based panelists, representing 93.3\% of the total panel
and 96.3\% of the total panel weights.

@@

When combining responses from different surveys, weights are rescaled
with respect to the total population, so that their order after
combining still reflects their relative importance. We \emph{combine}
(rather than cumulate) a set of \(\ell\) surveys \(S_1\), \(S_2\),
\ldots{} \(S_\ell\) (with \$ \ell \ge 2\$ ) as described in
O'Muircheartaigh and Pedlow (2002), by multiplying weights in \(S_i\) by
\(\lambda_i\), for \(1 \le i \le \ell\). \(\lambda_i \in [0,1]\) with
\(\sum_i \lambda_i = 1\) is given as

@@

The data for this paper were collected in several rounds. The resulting
number of participants, effective sample sizes, and \(\lambda\) values
are shown in \autoref{tbl-rounds}.

\hypertarget{tbl-rounds}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
\caption{\label{tbl-rounds}Survey rounds: dates, number of participants
(nominal sample size), effective sample size, sum of weights, and
factors for the combining of surveys as discussed above.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Date
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\# Participants
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Effective Sample Size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Sum of weights \(\sum_i w_i\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(\lambda_i\)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Date
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\# Participants
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Effective Sample Size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Sum of weights \(\sum_i w_i\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(\lambda_i\)
\end{minipage} \\
\midrule()
\endhead
Round 1 & April 2022 & 933 & 521.1 & 934.9 & \(\lambda_1\) = 0.343 \\
Round 2 & May 2022 & 953 & 485.7 & 953.4 & \(\lambda_2\) = 0.320 \\
Round 3 & June 2022 & 921 & 513.5 & 923.1 & \(\lambda_3\) = 0.338 \\
\textbf{Total} & --- & 2807 & 1520.8 & & \\
\bottomrule()
\end{longtable}

@@

\caption{\label{fig-alignment}On the left (a), a stacked barchart shows
the number of respondents with correct (green) and incorrect (grey)
responses to the two comparison questions. When tiles are aligned along
the same axis, more than twice the number of responses are correct. The
shaded area along the top of the green bars corresponds to 95\%
confidence intervals around (marginal) correct responses. On the right
(b), a parallel coordinate plot shows all combinations of responses.
There's a huge asymmetry in the number of responses where participants
answered only one of the questions correctly.}

@@

where \(\ell \in \{1, 2\}\); \(i \in \{1, 2\}\) is the comparison type
(1 = Aligned, 2 = Unaligned), and \(j \in \{1, 2, 3\}\) is the chart
design, with 1 = Vertical, 2 = Horizontal, and 3 = Horizontal wide. The
estimated values and 95\% confidence intervals are shown in
Table~\ref{tbl-rounds-123}. The lowercase letters indicate significances
between pairs of estimates: two estimates within the same column are
significantly different at 5\% level if they do not have any letter in
common (Piepho 2004).

\hypertarget{tbl-rounds-123}{}
\begin{longtable}{lrrlrrl}
\caption{\label{tbl-rounds-123}Estimated odds from the cell-means model for response patterns. Letters
behind numbers indicate pairwise significances. Within the same column,
values are significantly different (at 5\%) if they do not share the
same letter. }\tabularnewline

\caption*{
{\large Odds of selecting responses by task and chart type}
} \\ 
\toprule
 & \multicolumn{3}{c}{correct | same or wrong} & \multicolumn{3}{c}{correct or same | wrong} \\ 
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
 & Est. &           [95\% CI] &   & Est. &           [95\% CI] &   \\ 
\midrule
\multicolumn{7}{l}{Unaligned} \\ 
\midrule
Horizontal & $0.22$ &  [$0.15$, $0.32$] & a & $5.54$ &  [$4.04$, $7.59$] & a \\ 
Horizontal wide & $0.26$ &  [$0.19$, $0.37$] & ab & $6.20$ &  [$4.46$, $8.61$] & a \\ 
Vertical & $0.41$ &  [$0.36$, $0.47$] & b & $6.90$ &  [$5.75$, $8.28$] & a \\ 
\midrule
\multicolumn{7}{l}{Aligned} \\ 
\midrule
Horizontal & $0.63$ &  [$0.49$, $0.81$] & c & $14.02$ &  [$9.23$, $21.30$] & b \\ 
Horizontal wide & $2.67$ &  [$2.04$, $3.48$] & d & $17.12$ &  [$10.49$, $27.95$] & b \\ 
Vertical & $1.51$ &  [$1.33$, $1.71$] & e & $11.33$ &  [$8.99$, $14.28$] & b \\ 
\bottomrule
\end{longtable}

@@

Using linear scores for the response of certainty, with `not certain at
all' assigned a score of 1 and `extremely certain' assigned a score of
5, we can estimate the effects of task, chart design, and correctness on
certainty by using a cell-means model of the form:

@@

where \(k = 1, ..., N\), \(\mu_{ij\ell(k)}\) is average certainty
(measured on a scale from 1 to 5) of the four combinations of task and
correctness by each of the three designs, where \(i = 1, 2\) encodes
unaligned/aligned, and \(j=1,2\) encodes wrong, correct, and
\(\ell = 1, 2, 3\) encodes horizontal, wide horizontal, and vertical
stacked bar charts, respectively. We also assume that errors are
normally distributed,
i.e.~\(\epsilon_k \stackrel{i.i.d}{\sim} N(0, \sigma^2)\) for all
\(k = 1, ..., N\). The results are shown in Figure~\ref{fig-certainty}.
We find that the highest scores for certainty are associated with the
aligned task, but certainty scores are not significantly different
between correct and wrong responses. Scores are (mostly) significantly
lower for the unaligned task. Interestingly, the lowest certainty scores
are associated with correct responses to the unaligned task; although
respondents successfully complete the task, they are less certain in
their selection.

@@

\caption{\label{fig-certainty}Certainty by task, chart design, and
correctness. Correct answers in the unaligned task have the lowest
certainty scores associated with them. Incorrect responses on the
unaligned task show a significant boost in certainty in the vertical
stacked bar. For horizontal stacked bars the difference between aligned
and unaligned charts matters significantly. Error bars around the points
show estimated scores (on a scale of 1 to 5) with corresponding 95\% CI.
The letters underneath the bars indicate significance at 5\%. The scores
for two bars are significantly different if they do not share a letter.}

@@

\caption{\label{fig-demographics}Response patterns by alignment and
demographic levels. For aligned tiles, gender, age, and education are
not significant factors. However, income levels do have a (small)
effect. When income levels increase, the percentage of wrong answers
decreases, while the percentage of correct answers increases slightly
(significant at 5\%). For the more difficult task of comparing unaligned
tiles, higher levels of education and higher levels of income are
associated with a significant increase of panelists choosing a response
of `they are the same', resulting in significant decreases of both
correct answers and wrong answers.}
